\bf The forestecology R package for fitting and assessing models of interspecies competitive effects on the growth of trees

### 2021-03-03 paper/forestecology.Rmd
convert all sp to factors internally

We load both 2008 and 2014 SCBI census data `.csv` files as they existed on GitHub on November 20, 2020. After selecting only the relevant variables, we perform a few additional data wrangling steps: convert the character variable with the date of measurement to be of explicit type `date`, convert DBH to be in centimeters^[A rule of thumb to determine the units of DBH is to check if the smallest non-zero and non-missing measurement is 1 or 10. If the former, then centimeters. If the later, then millimeters. This is because ForestGEO protocols state that only trees with DBH greater or equal to 1cm should be included in censuses.], convert the `sp` variable containing species information from type `chr` character to `fct` factor.^[In our spatial cross-validation algorithm in Section \ref{spatial-cross-validation} issues can occur when rare species do not occur in the training set, but then are encountered in the test set. This risk is mitigated by representing `sp` as a factor variable, which has a complete list of all levels of the categorical variable.] Furthermore, in order to speed up computation for purposes of this example, we only consider a 9 ha subsection of the 25.6 ha of the SCBI site: `gx` from 0--300 instead of 0--400 and `gy` from 300--600 instead of 0--640.

```{r scbi-load-data}
census_2013_scbi <- read_csv("scbi.stem2.csv") %>%
  select(stemID, sp, date = ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = mdy(date),
    dbh = as.numeric(dbh)/10,
    sp = factor(sp)
  ) %>%
  filter(gx < 300, between(gy, 300, 600))

census_2018_scbi <- read_csv("scbi.stem3.csv") %>%
  select(stemID, sp, date = ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = mdy(date),
    dbh = as.numeric(dbh)/10,
    sp = factor(sp)
  ) %>%
  filter(gx < 300, between(gy, 300, 600))
```

@@

The output `growth_scbi` is a single data frame of class `sf` that includes variables `growth`, the average annual growth in DBH (cm y^-1^) for all stems that were alive at both time points, and `geometry`, the `sf` package's encoding of geolocations of type `<POINT>`. Furthermore, the variables that should remain unchanged between censuses appear only once, such as location variables `gx` and `gy`; as well as species-related variables. Variables that should change between censuses are suffixed with `1` and `2` indicating the earlier and later censuses, such as `dbh1/dbh2` and `codes1/codes2`. Site data that does not align with this convention will need to be transformed for use with the `compute_growth()` function. However, in the end, all that matters is that the growth of all stems is saved in a data frame of class `sf` and, at a minimum, the data contains the variables above. TODO: enumerate what these variables are and say if they need exact names

>>>>

We load both 2008 and 2014 SCBI census data `.csv` files as they existed on GitHub on November 20, 2020. After selecting only the relevant variables, we perform a few additional data wrangling steps: convert the character variable with the date of measurement to be of explicit type `date`, convert DBH to be in centimeters^[A rule of thumb to determine the units of DBH is to check if the smallest non-zero and non-missing measurement is 1 or 10. If the former, then centimeters. If the later, then millimeters. This is because ForestGEO protocols state that only trees with DBH greater or equal to 1cm should be included in censuses.] Furthermore, in order to speed up computation for purposes of this example, we only consider a 9 ha subsection of the 25.6 ha of the SCBI site: `gx` from 0--300 instead of 0--400 and `gy` from 300--600 instead of 0--640.

```{r scbi-load-data}
census_2013_scbi <- read_csv("scbi.stem2.csv") %>%
  select(stemID, sp, date = ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = mdy(date),
    dbh = as.numeric(dbh)/10
  ) %>%
  filter(gx < 300, between(gy, 300, 600))

census_2018_scbi <- read_csv("scbi.stem3.csv") %>%
  select(stemID, sp, date = ExactDate, gx, gy, dbh, codes, status) %>%
  mutate(
    date = mdy(date),
    dbh = as.numeric(dbh)/10
  ) %>%
  filter(gx < 300, between(gy, 300, 600))
```

@@

The output `growth_scbi` is a single data frame of class `sf` that includes variables `growth`, the average annual growth in DBH (cm y^-1^) for all stems that were alive at both time points, and `geometry`, the `sf` package's encoding of geolocations of type `<POINT>`. Furthermore, the variables that should remain unchanged between censuses appear only once, such as location variables `gx` and `gy`; as well as species-related variables. Variables that should change between censuses are suffixed with `1` and `2` indicating the earlier and later censuses, such as `dbh1/dbh2` and `codes1/codes2`. Site data that does not align with this convention will need to be transformed for use with the `compute_growth()` function. However, in the end, all that matters is that the growth of all stems is saved in a data frame of class `sf` and, at a minimum, the data contains the variables above. Lastly, the species variable `sp` is automatically converted to a factor^[In our spatial cross-validation algorithm in Section \ref{spatial-cross-validation} issues can occur when rare species do not occur in the training set, but then are encountered in the test set. This risk is mitigated by representing `sp` as a factor variable, which has a complete list of all levels of the categorical variable.].

TODO: enumerate what these variables are and say if they need exact names

### 2021-03-03 paper/forestecology.tex
convert all sp to factors internally

We load both 2008 and 2014 SCBI census data \texttt{.csv} files as they
existed on GitHub on November 20, 2020. After selecting only the
relevant variables, we perform a few additional data wrangling steps:
convert the character variable with the date of measurement to be of
explicit type \texttt{date}, convert DBH to be in centimeters\footnote{A
  rule of thumb to determine the units of DBH is to check if the
  smallest non-zero and non-missing measurement is 1 or 10. If the
  former, then centimeters. If the later, then millimeters. This is
  because ForestGEO protocols state that only trees with DBH greater or
  equal to 1cm should be included in censuses.}, convert the \texttt{sp}
variable containing species information from type \texttt{chr} character
to \texttt{fct} factor.\footnote{In our spatial cross-validation
  algorithm in Section \ref{spatial-cross-validation} issues can occur
  when rare species do not occur in the training set, but then are
  encountered in the test set. This risk is mitigated by representing
  \texttt{sp} as a factor variable, which has a complete list of all
  levels of the categorical variable.} Furthermore, in order to speed up
computation for purposes of this example, we only consider a 9 ha
subsection of the 25.6 ha of the SCBI site: \texttt{gx} from 0--300
instead of 0--400 and \texttt{gy} from 300--600 instead of 0--640.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{census_}\DecValTok{2013}\NormalTok{_scbi <-}\StringTok{ }\KeywordTok{read_csv}\NormalTok{(}\StringTok{"scbi.stem2.csv"}\NormalTok{) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{select}\NormalTok{(stemID, sp, }\DataTypeTok{date =}\NormalTok{ ExactDate, gx, gy, dbh, codes, status) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}
    \DataTypeTok{date =} \KeywordTok{mdy}\NormalTok{(date),}
    \DataTypeTok{dbh =} \KeywordTok{as.numeric}\NormalTok{(dbh)}\OperatorTok{/}\DecValTok{10}\NormalTok{,}
    \DataTypeTok{sp =} \KeywordTok{factor}\NormalTok{(sp)}
\NormalTok{  ) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{filter}\NormalTok{(gx }\OperatorTok{<}\StringTok{ }\DecValTok{300}\NormalTok{, }\KeywordTok{between}\NormalTok{(gy, }\DecValTok{300}\NormalTok{, }\DecValTok{600}\NormalTok{))}

\NormalTok{census_}\DecValTok{2018}\NormalTok{_scbi <-}\StringTok{ }\KeywordTok{read_csv}\NormalTok{(}\StringTok{"scbi.stem3.csv"}\NormalTok{) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{select}\NormalTok{(stemID, sp, }\DataTypeTok{date =}\NormalTok{ ExactDate, gx, gy, dbh, codes, status) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}
    \DataTypeTok{date =} \KeywordTok{mdy}\NormalTok{(date),}
    \DataTypeTok{dbh =} \KeywordTok{as.numeric}\NormalTok{(dbh)}\OperatorTok{/}\DecValTok{10}\NormalTok{,}
    \DataTypeTok{sp =} \KeywordTok{factor}\NormalTok{(sp)}
\NormalTok{  ) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{filter}\NormalTok{(gx }\OperatorTok{<}\StringTok{ }\DecValTok{300}\NormalTok{, }\KeywordTok{between}\NormalTok{(gy, }\DecValTok{300}\NormalTok{, }\DecValTok{600}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

The output \texttt{growth\_scbi} is a single data frame of class
\texttt{sf} that includes variables \texttt{growth}, the average annual
growth in DBH (cm y\textsuperscript{-1}) for all stems that were alive
at both time points, and \texttt{geometry}, the \texttt{sf} package's
encoding of geolocations of type
\texttt{\textless{}POINT\textgreater{}}. Furthermore, the variables that
should remain unchanged between censuses appear only once, such as
location variables \texttt{gx} and \texttt{gy}; as well as
species-related variables. Variables that should change between censuses
are suffixed with \texttt{1} and \texttt{2} indicating the earlier and
later censuses, such as \texttt{dbh1/dbh2} and \texttt{codes1/codes2}.
Site data that does not align with this convention will need to be
transformed for use with the \texttt{compute\_growth()} function.
However, in the end, all that matters is that the growth of all stems is
saved in a data frame of class \texttt{sf} and, at a minimum, the data
contains the variables above. TODO: enumerate what these variables are
and say if they need exact names

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# }\AlertTok{TODO}\CommentTok{: Below reconcile the number of rows as they off by one from growth_scbi %>% }
\CommentTok{# filter(!is.na(growth) & !buffer). Perhaps by removing NA's in the growth_scbi stage?}
\NormalTok{focal_vs_comp_scbi <-}\StringTok{ }\NormalTok{growth_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{create_focal_vs_comp}\NormalTok{(comp_dist, }\DataTypeTok{cv_grid_sf =}\NormalTok{ blocks_scbi, }\DataTypeTok{id =} \StringTok{"stemID"}\NormalTok{)}
\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{select}\NormalTok{(}\OperatorTok{-}\NormalTok{foldID)}
\CommentTok{## # A tibble: 6,296 x 6}
\CommentTok{##   focal_ID focal_sp   dbh   geometry growth comp             }
\CommentTok{##      <dbl> <fct>    <dbl>    <POINT>  <dbl> <list>           }
\CommentTok{## 1        4 nysy     13.6  (14.2 428)  0.103 <tibble [20 x 4]>}
\CommentTok{## 2        5 havi      8.8   (9.4 436)  0.150 <tibble [32 x 4]>}
\CommentTok{## 3       79 tiam     47.7    (40 381) -0.161 <tibble [20 x 4]>}
\CommentTok{## 4       80 caca      5.15 (38.7 422)  0.253 <tibble [12 x 4]>}
\CommentTok{## 5       96 libe      2.3    (60 310)  0.262 <tibble [14 x 4]>}
\CommentTok{## # ... with 6,291 more rows}
\end{Highlighting}
\end{Shaded}

>>>>

We load both 2008 and 2014 SCBI census data \texttt{.csv} files as they
existed on GitHub on November 20, 2020. After selecting only the
relevant variables, we perform a few additional data wrangling steps:
convert the character variable with the date of measurement to be of
explicit type \texttt{date}, convert DBH to be in centimeters\footnote{A
  rule of thumb to determine the units of DBH is to check if the
  smallest non-zero and non-missing measurement is 1 or 10. If the
  former, then centimeters. If the later, then millimeters. This is
  because ForestGEO protocols state that only trees with DBH greater or
  equal to 1cm should be included in censuses.} Furthermore, in order to
speed up computation for purposes of this example, we only consider a 9
ha subsection of the 25.6 ha of the SCBI site: \texttt{gx} from 0--300
instead of 0--400 and \texttt{gy} from 300--600 instead of 0--640.

\begin{Shaded}
\begin{Highlighting}[]
\NormalTok{census_}\DecValTok{2013}\NormalTok{_scbi <-}\StringTok{ }\KeywordTok{read_csv}\NormalTok{(}\StringTok{"scbi.stem2.csv"}\NormalTok{) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{select}\NormalTok{(stemID, sp, }\DataTypeTok{date =}\NormalTok{ ExactDate, gx, gy, dbh, codes, status) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}
    \DataTypeTok{date =} \KeywordTok{mdy}\NormalTok{(date),}
    \DataTypeTok{dbh =} \KeywordTok{as.numeric}\NormalTok{(dbh)}\OperatorTok{/}\DecValTok{10}
\NormalTok{  ) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{filter}\NormalTok{(gx }\OperatorTok{<}\StringTok{ }\DecValTok{300}\NormalTok{, }\KeywordTok{between}\NormalTok{(gy, }\DecValTok{300}\NormalTok{, }\DecValTok{600}\NormalTok{))}

\NormalTok{census_}\DecValTok{2018}\NormalTok{_scbi <-}\StringTok{ }\KeywordTok{read_csv}\NormalTok{(}\StringTok{"scbi.stem3.csv"}\NormalTok{) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{select}\NormalTok{(stemID, sp, }\DataTypeTok{date =}\NormalTok{ ExactDate, gx, gy, dbh, codes, status) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{mutate}\NormalTok{(}
    \DataTypeTok{date =} \KeywordTok{mdy}\NormalTok{(date),}
    \DataTypeTok{dbh =} \KeywordTok{as.numeric}\NormalTok{(dbh)}\OperatorTok{/}\DecValTok{10}
\NormalTok{  ) }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{filter}\NormalTok{(gx }\OperatorTok{<}\StringTok{ }\DecValTok{300}\NormalTok{, }\KeywordTok{between}\NormalTok{(gy, }\DecValTok{300}\NormalTok{, }\DecValTok{600}\NormalTok{))}
\end{Highlighting}
\end{Shaded}

@@

The output \texttt{growth\_scbi} is a single data frame of class
\texttt{sf} that includes variables \texttt{growth}, the average annual
growth in DBH (cm y\textsuperscript{-1}) for all stems that were alive
at both time points, and \texttt{geometry}, the \texttt{sf} package's
encoding of geolocations of type
\texttt{\textless{}POINT\textgreater{}}. Furthermore, the variables that
should remain unchanged between censuses appear only once, such as
location variables \texttt{gx} and \texttt{gy}; as well as
species-related variables. Variables that should change between censuses
are suffixed with \texttt{1} and \texttt{2} indicating the earlier and
later censuses, such as \texttt{dbh1/dbh2} and \texttt{codes1/codes2}.
Site data that does not align with this convention will need to be
transformed for use with the \texttt{compute\_growth()} function.
However, in the end, all that matters is that the growth of all stems is
saved in a data frame of class \texttt{sf} and, at a minimum, the data
contains the variables above. Lastly, the species variable \texttt{sp}
is automatically converted to a factor\footnote{In our spatial
  cross-validation algorithm in Section \ref{spatial-cross-validation}
  issues can occur when rare species do not occur in the training set,
  but then are encountered in the test set. This risk is mitigated by
  representing \texttt{sp} as a factor variable, which has a complete
  list of all levels of the categorical variable.}.

TODO: enumerate what these variables are and say if they need exact
names

@@

\begin{Shaded}
\begin{Highlighting}[]
\CommentTok{# }\AlertTok{TODO}\CommentTok{: Below reconcile the number of rows as they off by one from growth_scbi %>% }
\CommentTok{# filter(!is.na(growth) & !buffer). Perhaps by removing NA's in the growth_scbi stage?}
\NormalTok{focal_vs_comp_scbi <-}\StringTok{ }\NormalTok{growth_scbi }\OperatorTok{%>%}
\StringTok{  }\KeywordTok{create_focal_vs_comp}\NormalTok{(comp_dist, }\DataTypeTok{blocks =}\NormalTok{ blocks_scbi, }\DataTypeTok{id =} \StringTok{"stemID"}\NormalTok{)}
\NormalTok{focal_vs_comp_scbi }\OperatorTok{%>%}\StringTok{ }
\StringTok{  }\KeywordTok{select}\NormalTok{(}\OperatorTok{-}\NormalTok{foldID)}
\CommentTok{## # A tibble: 6,296 x 6}
\CommentTok{##   focal_ID focal_sp   dbh   geometry growth comp             }
\CommentTok{##      <dbl> <fct>    <dbl>    <POINT>  <dbl> <list>           }
\CommentTok{## 1        4 nysy     13.6  (14.2 428)  0.103 <tibble [20 x 4]>}
\CommentTok{## 2        5 havi      8.8   (9.4 436)  0.150 <tibble [32 x 4]>}
\CommentTok{## 3       79 tiam     47.7    (40 381) -0.161 <tibble [20 x 4]>}
\CommentTok{## 4       80 caca      5.15 (38.7 422)  0.253 <tibble [12 x 4]>}
\CommentTok{## 5       96 libe      2.3    (60 310)  0.262 <tibble [14 x 4]>}
\CommentTok{## # ... with 6,291 more rows}
\end{Highlighting}
\end{Shaded}
